\begin{abstract}
This thesis presents a search for light charged Higgs bosons produced in top quark decays using proton-proton collision data collected by the CMS experiment at the Large Hadron Collider. The data correspond to an integrated luminosity of 138\fbinv recorded during Run~2 (2016--2018) at a center-of-mass energy of $\sqrt{s} = 13~\TeV$.

The search targets charged Higgs bosons produced in top quark decays, $\PQt \to \PHc\PQb$, with subsequent decays $\PHc \to \PW\PA$ and $\PA \to \mu^{+}\mu^{-}$, where \PHc is the charged Higgs boson and \PA is the CP-odd pseudoscalar predicted by Two-Higgs-Doublet Model (2HDM) extensions of the Standard Model. The final state consists of three charged leptons (one electron and two muons, or three muons), missing transverse momentum from neutrinos, and at least two jets from bottom quark hadronization.

A key advancement of this analysis compared to previous searches is the inclusion of off-shell $\PHc \to \PW\PA$ decays, extending coverage to mass configurations where the decay proceeds through a virtual \PW boson. The search covers charged Higgs boson masses from 70 to 160\GeV and pseudoscalar masses from 15\GeV up to $\mHc - 5\GeV$.

A ParticleNet-based graph neural network classifier is employed to discriminate signal events from \PZ boson associated backgrounds in the ``on-Z'' mass region where the signal dimuon mass peak overlaps with the \PZ resonance. Signal extraction is performed through a binned maximum likelihood fit to the dimuon invariant mass distribution using the asymptotic CL$_s$ method.

No statistically significant excess above the Standard Model prediction is observed. Upper limits at 95\% confidence level are set on the signal branching ratio, providing constraints on Two-Higgs-Doublet Model scenarios, particularly the Type-I 2HDM at large \tanb where the bosonic decay mode $\PHc \to \PW\PA$ can become dominant.

This result represents the first search for the decay $\PHc \to \PW\PA$ including off-shell effects, providing comprehensive coverage of the 2HDM parameter space accessible at the LHC for light charged Higgs bosons.
\end{abstract}
